Shallow EDSLs and Object-Oriented Programming

### 2016-06-17 paper/paper.tex
more text

This paper argues that OOP abstractions, including \emph{inheritance}
and \emph{subtyping}, increase the modularity and reuse of shallow
DSLs when compared to classical procedural abstraction. We make this
argument by taking a recent paper by Gibbons and Wu, where procedural
abstraction is used in Haskell to model a simple EDSL, and we recode
that EDSL in Java. Although from the \emph{syntactical} point of view
there are obvious disadvantages in the Java version, from the semantic
and modularity point of view the Java version has clear advantages.
To alleviate some of the syntactical disadvantages of Java, we create
an annotation inspired by \emph{family polymorphism} and a recent
solution to the Expression Problem.
The annotation uses transparent code generation techniques to
automatically eliminate large portions of boilerplate code.
To further illustrate the applicability of our tool and techniques, we conduct
several case studies using larger DSLs from the literature.
\end{abstract}

>>>>

This paper argues that OOP abstractions, including \emph{inheritance}
and \emph{subtyping}, increase the modularity and reuse of shallow
DSLs when compared to classical procedural abstraction. We make this
argument by taking a recent paper by Gibbons and Wu, where procedural
abstraction is used in Haskell to model a simple EDSL, and we recode
that EDSL in Java. Although from the \emph{syntactical} point of view
there are obvious disadvantages in the Java version, from the semantic
and modularity point of view the Java version has clear advantages.
%To alleviate some of the syntactical disadvantages of Java, we create
%an annotation inspired by \emph{family polymorphism} and a recent
%solution to the Expression Problem.
%The annotation uses transparent code generation techniques to
%automatically eliminate large portions of boilerplate code.
%To further illustrate the applicability of our tool and techniques, we conduct
%several case studies using larger DSLs from the literature.
\end{abstract}

### 2016-06-17 paper/sections/Introduction.tex
more text

Since Hudak's seminal paper on Embedded DSLs (EDSLs)~\cite{}, existing
languages (such as Haskell) have been used to directly encode DSLs. Two common approaches to EDSLs are the
so-called \emph{shallow} and \emph{deep} embeddings. The difference 
between these two styles of embeddings is commonly described as 
follows:

@@

This definition seems to be widely accepted and similar definitions
appear in many works~\cite{}. We believe that this definition is a
vague and not very accurate, often leading to some contradicting
claims. For example, in their work on EDSLs~\cite{}, Gibbons and Wu
claim that deep embeddings (which encode ASTs using algebraic
datatypes in Haskell) allow adding new DSL interpretations easily, but
they make adding new language constructs difficult. In contrast
Gibbons and Wu claim that shallow embeddings have dual modularity
properties: new cases are easy to add, but new interpretations are
hard.  A contradiction happens if instead of using Haskell and
algebraic datatypes, one uses an OO language to encode an AST, for
example with the {\sc Composite} pattern. In that case, according to
definition of deep embedding above, the OO approach classifies as a
deep embedding. Nevertheless, as the Expression Problem~\cite{} tell
us, in the OO approach adding new cases is easy, but adding
interpretations is hard. Thus we have a contradiction, since we have
an AST representation (i.e. a deep embedding) with the modularity
properties of shallow embeddings.

We believe that the core of problem is that ASTs can be represented 
in multiple ways. In particular, it is well know that functions alone 
are enough to encode datastructures such as ASTs (via Church
encodings~\cite{}). Distinguishing deep and shallow
embeddings solely on whether a datastructure is being used 
or whether functions are used to encode interpretations directly 
falls short when the functions themselfves represent an AST.

In this paper we propose a different definition for deep and shallow
embeddings, by distinguishing on what kind of data abstraction is used
by each encoding. Deep embeddings use user-defined types
(i.e. disjoint unions or algebraic datatypes), whereas shallow
embeddings use procedural abstraction. This distinction is inspired by
Reynolds 1975 paper~\cite{} on user-defined types and procedural
abstraction being complementary approaches to data abstraction, which
many view~\cite{} as the first documented discussion on the Expression
Problem.

Shallow Embedded Domain Specific Languages (EDSLs) use
\emph{procedural abstraction} to directly encode a Domain Specific
Language into an existing host language. Procedural abstraction has
been argued to be the essence of Object-Oriented Programming (OOP). Given
that Object-Oriented Languages have evolved over more than 50 years 
to improve the use of procedural abstraction, they ought to have some 
advantages to encode shallow EDSLs.

This paper argues that OOP abstractions, including \emph{inheritance}
and \emph{subtyping}, increase the modularity and reuse of shallow
DSLs when compared to classical procedural abstraction. We make this
argument by taking a recent paper by Gibbons and Wu, where procedural
abstraction is used in Haskell to model a simple EDSL, and we recode
that EDSL in Java. Although from the \emph{syntactical} point of view
there are obvious disadvantages in the Java version, from the semantic
and modularity point of view the Java version has clear advantages.
To alleviate some of the syntactical disadvantages of Java, we create
an annotation inspired by \emph{family polymorphism} and a recent
solution to the Expression Problem. 
The annotation uses transparent code generation techniques to
automatically eliminate large portions of boilerplate code.
To further illustrate the applicability of our tool and techniques, we conduct 
several case studies using larger DSLs from the literature.

>>>>

Since Hudak's seminal paper on Embedded DSLs (EDSLs)~\cite{}, existing
languages (such as Haskell) have been used to directly encode
DSLs. Two common approaches to EDSLs are the so-called \emph{shallow}
and \emph{deep} embeddings. The origin of that terminology can be
attributed to Boulton's work~\cite{}. The difference between these
two styles of embeddings is commonly described as follows:

@@

%This definition is widely accepted and similar definitions appear in
%many other works~\cite{}. 
We argue that this definition is vague,
and often leads to some contradicting claims. For example, in their
work on EDSLs~\cite{}, Gibbons and Wu claim that deep embeddings
(which encode ASTs using algebraic datatypes in Haskell) allow adding
new DSL interpretations easily, but they make adding new language
constructs difficult. In contrast Gibbons and Wu claim that shallow
embeddings have dual modularity properties: new cases are easy to add,
but new interpretations are hard.  A contradiction happens if, instead
of using Haskell and algebraic datatypes, one uses an OO language to
encode an AST, for example with the {\sc Composite} pattern. In that
case, \emph{according to definition of deep embedding above, the OO
  approach classifies as a deep embedding}. Nevertheless, as the
Expression Problem~\cite{} tell us, in the OO approach adding new
language constructs is easy, but adding interpretations is hard. Thus
we have a contradiction, since we have an AST representation (i.e. a
deep embedding) with the modularity properties of shallow embeddings.

We believe that the core of problem is that ASTs can be represented in
multiple ways. In particular, it is well know that functions alone are
enough to encode datastructures such as ASTs (via Church
encodings~\cite{}).  Distinguishing deep and shallow embeddings based
solely on whether a ``real'' datastructure is being used or not is
misleading.  Moreover, it gives the impression that shallow embeddings
are significantly less expressive than deep embeddings, because they
do not have access to the datastructure.
Gibbons and Wu themselves feel uneasy with the definition of shallow 
embeddings when they say:
``\emph{So it turns out that the syntax of the DSL is not really as ephemeral
in a shallow embedding as Boulton's choice of terms suggests.}''

A first goal in this paper is to propose distinguishing EDSLs in terms of the data
abstraction used to model the language constructs instead.  We follow
Reynold's classification~\cite{} of data abstractions:
\emph{procedural abstraction} and \emph{user-defined types}. It is
clear that shallow embeddings use \emph{procedural
  abstraction}~\cite{}: the DSLs are modelled by interpretation
functions. Therefore, the other implementation option for EDSLs is to use
\emph{user-defined types}. In Reynolds terminology user-defined types
mean disjoint union types, which are nowadays commonly available in
modern languages as \emph{algebraic datatypes}. Disjoint union types 
can also be emulated in OOP using the {\sc Visitor} pattern. A
distinction based on data abstraction provides a remedy for Gibbons 
and Wu's contradiction. An EDSL implemented with algebraic
datatypes falls into the category of user-defined types, 
while a Composite-based OO implementation falls under procedural
abstraction. For the rest of the paper we identify shallow EDSLs 
with EDSLs implemented using procedural abstraction.

The second and main goal of this paper is to show that OOP languages have
advantages for the implementation of shallow embeddings.
As Cook~\cite{} argued procedural abstraction is the essence of
Object-Oriented Programming. If we accept Cook's view, the
implementation of a shallow DSL in OOP languages should simply
correspond to a standard object-oriented program. Given
that Object-Oriented Languages have evolved over more than 50 years 
to improve the use of procedural abstraction, they ought to have some 
advantages to encode shallow EDSLs.

We show that OOP abstractions, including \emph{inheritance}
and \emph{subtyping}, increase the modularity and reuse of shallow
DSLs when compared to classical procedural abstraction. We make this
argument by taking a recent paper by Gibbons and Wu, where procedural
abstraction is used in Haskell to model a simple EDSL, and we recode
that EDSL in Java. Although from the \emph{syntactical} point of view
there are obvious disadvantages in the Java version, from the semantic
and modularity point of view the Java version has clear advantages.

This paper has two goals:

\begin{itemize}

\item To argue that

\end{itemize}
